3D Regression Heat Map Analysis of Population Study Data

### 2015-06-27 draft.tex
minor last revisions

\paragraph{Visual analysis of epidemiological data.}
%%
The work of Zhang et al.~\cite{Gotz2014, Zhang2014} is closest to ours regarding the application of a visual analysis of cohort study data.
%%
They present \emph{Cohort Analysis via Visual Analytics} (\emph{CAVA}), a framework that distinguishes three major elements of a cohort study data analysis: \emph{cohort data} (and its manipulation using operations), \emph{views} and \emph{analytics}.
%%
They use the system to find longitudinal pathways for diseases on the basis of health records.
%%
%Sub-groups are automatically created by incorporating \emph{analytics} and through selection in expert-guided \emph{views}.
%%
%They do not go into the detail w.r.t the analytics used, they seem to not incorporate confounder analysis (other than dividing groups by gender) and the user study lacks concrete goals.
%%
%Their data differs strongly from ours, having multiple moments for each subject, yielding a approach focused on finding pathways of interest.
%%
We incorporate their requirements, which consist of a \emph{flexible} and \emph{iterative analysis}. %, meaning that the system has to adapt to different requirements in different analysis steps.
%%
%We implement this by supporting both hypothesis-based and hypothesis-free analysis.
%%
Angelelli et al.~\cite{Angelelli} visualize image-derived an non-image data using cube data structures with focus on comparison and knowledge extraction.
%%
%Their proposed engine represents multiple heterogenous cohort study data sets as normalized data cubes.
%%
%Thus, data redundancy is avoided and runtime measure aggregation can be carried out when measures of the different cubes have to be combined.
%%
%This allows for saving memory space.
%%
%Their presented aggregation functions allow for a fast data calculation to map it using different linked visualizations, such as scatter plots, bar charts or histograms.
%%
They use \emph{Pearsson's} $r$ to characterize relationships with target features and employ list views and scatter plots to visualize and rank them.
%%
%We use regression models and associated $R^2$ values to depict relationships between a target feature and multiple independent features instead of one and employ the information into one global visualization.
%%
In contrast, we focus on a fast large scale correlation analysis incorporating many features to derive insights.
%%
%\paragraph{Visual Analysis in Public Health.}
%%--
% Shneiderman et al.~\cite{Shneiderman2013} highlight the necessity of interactive visualizations in healthcare data. %, e.g. using \emph{Health 2.0}, which is an umbrella for web-based health-informatics frameworks.
% %%
% %They divide Health 2.0 in three domains, \emph{Personal health information}, \emph{clinical health information}, and \emph{public health information}.
% %%
% %We focus on the latter.
% %%
% %Shneiderman and colleagues also present challenges,
% The presented challenges in this application require the systems to be sufficiently comprehensible and intuitive to fit into the short time span that clinicians have for analyzing data.
% %%
% Furthermore, the systems should provide comparative relationship visualization, characterization of similarity and presenting risks.
%%
%We use the challenges as design guidelines for our web-based prototype, focusing on functional clean designs to allow fast insight into relationships towards diseases.
%%

@@

As described in Section~\ref{sec:RegressionAnalysis}, regression analysis is the statistical tool of choice for analyzing relationships in epidemiological data.
%%
A regression model is based on expert knowledge.
%%
There is no rule how to apply models to a given set of features. Thus, they have to be applied with care.
%%
% \design{
% %%
% Key requirements based on the epidemiological application domain on the 3D Regression Heat Map are:
% %%
% \begin{itemize}
% 	\item Show relationships of features w.r.t. their explanatory power towards the target feature (e.g., a disease or unhealthy condition).
% 	\item Allow for modeling assumptions regarding the target feature.
% 	\item The tool has to be easily accessible and simultaneously secure the data its processing.
% \end{itemize}
% %%
% }
\subsection{Regression Heat Map Description Using Regression Formula Notation}
%%
Expert knowledge modeling is carried out using \emph{regression formulas}.
%Expert knowledge is integrated into the analysis using regression formulas.
%%
The formula input influences the type of the chosen regression method as well as the \emph{independent} features describing the target (recall Sect.~\ref{sec:RegressionAnalysis}).
%%

@@

\subsection{Analysis Workflow} \label{sec:Workflow}
%%
% \begin{figure}[htb]
%  \centering
%  \includegraphics[width=1.0\linewidth]{figures/workflow}
%  \caption{
%  Different workflow types using the \emph{3D Regression Heat Map}.
%  %%
%  [\textbf{1.AF}] The workflow starts by declaring a formula to specify a hypothesis, or to use a predefined formula for a hypothesis-free analysis.
%  %%
%  [\textbf{2.SI}] The \emph{3D Regression Heat Map} is then visualized.
%  %%
%  The user has the option to either adjust the formula, adjust the transfer function or to derive details-on-demand on models.
%  %%
%  [\textbf{3.ZF}] Insights into the data yield either an adjustment of the current formula or a selection of a difference view.
%  %%
%  The latter is used to compare \emph{3D Regression Heat Maps}.
%  %%
%  [\textbf{4.DD}] Details about features using the 2D heat map representation yield insights and hypotheses about feature relations.
%  }
%   \label{fig:Workflow}
% \end{figure}
%%
\noindent Our \emph{3D Regression Heat Map} model is well suited for different workflow analysis techniques, based on the Visual Analytics (VA) Mantra of Keim et al.~\cite{Keim}:

@@

Both data sets are unusual for epidemiological analysis regarding their feature extent.
%%
Usually, only a few features depicting a hypothesis are compiled into a data set to assess them using statistical tools.
%%
The herein used data sets comprise several hundred features.
%%
Our method focuses on data exploration and knowledge extraction and requires a wide scope of sociodemographic, medical and lifestyle features.
%%
\subsection{Participants, Setup and Procedure}
%%
The knowledge discovery capabilities of a system are difficult to measure.
%%
The \emph{Visual Data Analysis and Reasoning} (\emph{VDAR}) technique proposed by Lam et al.~\cite{Lam2012} is focused on the characterization of a system's ability to generate hypotheses and explore the data in order to extract information.
%%
\emph{VDAR} can be carried out based on case studies using thinking-aloud techniques to comprehend the user's reasoning and thought process.
%%
We employ \emph{VDAR} for analyzing our system.
%%
\paragraph{Participants, setup and procedure.}
%%
We conducted a web-based analysis by using an online meeting software, which features voice chat as well as screen sharing.
%%
Starting an analysis using these techniques takes about 5-10 minutes of setup time.
%%
The sessions started with an initial overview of the system, showcasing its features and functionality.
%%
Afterwards, the experts used the system on their own computers.
%%
The screen-sharing function was still used to observe the actions of the experts.
%%
All sessions were video-recorded to be processed later on.
%%
We conducted the analysis with three participants.
%%
\emph{KH}, a clinician (10 years of experience) with focus on epidemiological research, is the domain expert for the breast density data set.
%%
She is a radiologist responsible for the SHIP-MRI acquisition and also for the mammography analysis.
%%
The hepatic steatosis data set is analyzed by \emph{UN}, a data scientist responsible for prior analysis of the data.
%%
The third participant is \emph{TI}, a statistician with focus on epidemiology (8 years of experience), who assesses the statistical reliability of the tool and the underlying methods without a focus on a specific data set.
%%
\subsection{The Hepatic Steatosis Data Set}
We employ the data set used by Niemann et al.~\cite{Niemann2014} to identify predictive features w.r.t. the reversible hepatic steatosis disorder.
%%
The dichotomous target feature is derived from the liver fat concentration measured using MRI scans.
%%
Liver fat concentrations of no more than 10\% are mapped to the `negative' class; values greater than 10\% are mapped to the `positive' class to indicate absence or presence of the disease.
%%
The data set contains labels for 578 participants. %for which the MRI data was available to that point.
%%
The MRI scans for each subject are only available in SHIP-2. %, so the target feature is only present the last study wave.

%%
Apart from the target feature, the data set contains 199 features comprising sociodemographic features (gender, age), consumption behavior (e.g., alcohol and tobacco), laboratory data (e.g., sera concentrations), and two features depicting the liver ultrasound.
%%
The acquisition wave is denoted using the appendix; 85 features with appendix \emph{s0} denote their affiliation to SHIP-0 (first study moment), 50 features for \emph{s1} and 55 for \emph{s2}, alongside with 10 time-independent SNPs (DNA base pairs).
%%
Niemann et al.~\cite{Niemann2014} show different class distributions of liver fat concentrations of women and men.
%%
For women, an association between age and liver fat was identified.
%%
An appropriate cut-off value of 52 years, which is the approximate entry age for the menopause was set, yielding the most homogeneous class distribution within the resulting subsets.
%%
Based on these observations, we perform our analysis on three populations: \emph{males}, \emph{females (all ages)} and \emph{females older than 52 years}.
%%
\subsection{The Breast Density Data Set}
%%
The breast density data set was compiled to find associations between the parenchyma tissue proportion in the female breast compared to other features in the data.
%%
Breast density is denoted as the ratio between parenchyma and cellular connective tissue and has been shown to be associated with breast cancer.
%%
Studies describe a four to five times increased risk of getting breast cancer for participants with a breast density above 50\% \cite{Mccormack2006}.

>>>>

\paragraph{Visual analysis of epidemiological data.}
%%
The work of Zhang et al.~\cite{Gotz2014, Zhang2014} is closest to ours regarding the application of a visual analysis of population study data.
%%
They present \emph{Cohort Analysis via Visual Analytics} (\emph{CAVA}), a framework that distinguishes three major elements of a cohort study data analysis: \emph{cohort data} (and its manipulation using operations), \emph{views} and \emph{analytics}.
%%
They use the system to find longitudinal pathways for diseases on the basis of health records.
%%
%Sub-groups are automatically created by incorporating \emph{analytics} and through selection in expert-guided \emph{views}.
%%
%They do not go into the detail w.r.t the analytics used, they seem to not incorporate confounder analysis (other than dividing groups by gender) and the user study lacks concrete goals.
%%
%Their data differs strongly from ours, having multiple moments for each subject, yielding a approach focused on finding pathways of interest.
%%
We incorporate their requirements, which consist of a \emph{flexible} and \emph{iterative analysis}. %, meaning that the system has to adapt to different requirements in different analysis steps.
%%
%We implement this by supporting both hypothesis-based and hypothesis-free analysis.
%%
Angelelli et al.~\cite{Angelelli} visualize image-derived an non-image data using cube data structures with focus on comparison and knowledge extraction.
%%
%Their proposed engine represents multiple heterogenous cohort study data sets as normalized data cubes.
%%
%Thus, data redundancy is avoided and runtime measure aggregation can be carried out when measures of the different cubes have to be combined.
%%
%This allows for saving memory space.
%%
%Their presented aggregation functions allow for a fast data calculation to map it using different linked visualizations, such as scatter plots, bar charts or histograms.
%%
They use \emph{Pearsson's} $r$ to characterize relationships with target features and employ list views and scatter plots to visualize and rank them.
%%
%We use regression models and associated $R^2$ values to depict relationships between a target feature and multiple independent features instead of one and employ the information into one global visualization.
%%
In contrast, we focus on a fast large scale correlation analysis incorporating many features to derive insights.
%%
%\paragraph{Visual Analysis in Public Health.}
%%--
% Shneiderman et al.~\cite{Shneiderman2013} highlight the necessity of interactive visualizations in healthcare data. %, e.g. using \emph{Health 2.0}, which is an umbrella for web-based health-informatics frameworks.
% %%
% %They divide Health 2.0 in three domains, \emph{Personal health information}, \emph{clinical health information}, and \emph{public health information}.
% %%
% %We focus on the latter.
% %%
% %Shneiderman and colleagues also present challenges,
% The presented challenges in this application require the systems to be sufficiently comprehensible and intuitive to fit into the short time span that clinicians have for analyzing data.
% %%
% Furthermore, the systems should provide comparative relationship visualization, characterization of similarity and presenting risks.
%%
%We use the challenges as design guidelines for our web-based prototype, focusing on functional clean designs to allow fast insight into relationships towards diseases.
%%

@@

Regression analysis is the statistical tool of choice for analyzing relationships in epidemiological data (recall Sect.~\ref{sec:RegressionAnalysis}).
%%
A regression model is based on expert knowledge.
%%
There is no rule how to apply models to a given set of features. Thus, they have to be applied with care.
%%
% \design{
% %%
% Key requirements based on the epidemiological application domain on the 3D Regression Heat Map are:
% %%
% \begin{itemize}
% 	\item Show relationships of features w.r.t. their explanatory power towards the target feature (e.g., a disease or unhealthy condition).
% 	\item Allow for modeling assumptions regarding the target feature.
% 	\item The tool has to be easily accessible and simultaneously secure the data its processing.
% \end{itemize}
% %%
% }
\subsection{Regression Heat Map Description Using Regression Formula Notation}
%%
Expert knowledge modeling is carried out using \emph{regression formulas}.
%Expert knowledge is integrated into the analysis using regression formulas.
%%
The formula input influences the type of the chosen regression method as well as the \emph{independent} features describing the target (recall Sect.~\ref{sec:RegressionAnalysis}).
%%

@@

\subsection{Analysis Workflow} \label{sec:Workflow}
%%
% \begin{figure}[htb]
%  \centering
%  \includegraphics[width=1.0\linewidth]{figures/workflow}
%  \caption{
%  Different workflow types using the \emph{3D Regression Heat Map}.
%  %%
%  [\textbf{1.AF}] The workflow starts by declaring a formula to specify a hypothesis, or to use a predefined formula for a hypothesis-free analysis.
%  %%
%  [\textbf{2.SI}] The \emph{3D Regression Heat Map} is then visualized.
%  %%
%  The user has the option to either adjust the formula, adjust the transfer function or to derive details-on-demand on models.
%  %%
%  [\textbf{3.ZF}] Insights into the data yield either an adjustment of the current formula or a selection of a difference view.
%  %%
%  The latter is used to compare \emph{3D Regression Heat Maps}.
%  %%
%  [\textbf{4.DD}] Details about features using the 2D heat map representation yield insights and hypotheses about feature relations.
%  }
%   \label{fig:Workflow}
% \end{figure}
%%
\noindent Our \emph{3D Regression Heat Map} is well suited for different workflow analysis techniques, based on the Visual Analytics (VA) Mantra of Keim et al.~\cite{Keim}:

@@

Both data sets are unusual for epidemiological analysis regarding their feature extent.
%%
Usually, only a few features depicting a hypothesis are compiled into a data set to assess them using statistical tools.
%%
The herein used data sets comprise several hundred features.
%%
Our method focuses on data exploration and knowledge extraction and requires a wide scope of sociodemographic, medical and lifestyle features.
%%
\subsection{Participants, Setup and Procedure}
%%
The knowledge discovery capabilities of a system are difficult to measure.
%%
The \emph{Visual Data Analysis and Reasoning} (\emph{VDAR}) technique proposed by Lam et al.~\cite{Lam2012} is focused on the characterization of a system's ability to generate hypotheses and explore the data in order to extract information.
%%
\emph{VDAR} can be carried out based on case studies using thinking-aloud techniques to comprehend the user's reasoning and thought process.
%%
We employ \emph{VDAR} for analyzing our system.
%%
\paragraph{Participants, setup and procedure.}
%%
We conducted a web-based analysis by using an online meeting software, which features voice chat as well as screen sharing.
%%
Starting an analysis using these techniques took about 5-10 minutes of setup time.
%%
The sessions started with an initial overview of the system, showcasing its features and functionality.
%%
Afterwards, the experts used the system on their own computers.
%%
The screen-sharing function was still used to observe the actions of the experts.
%%
All sessions were video-recorded to be processed later on.
%%
We conducted the analysis with three participants.
%%
\emph{KH}, a clinician (10 years of experience) with focus on epidemiological research, is the domain expert for the breast density data set.
%%
She is a radiologist responsible for the SHIP-MRI acquisition and also for the mammography analysis.
%%
The hepatic steatosis data set is analyzed by \emph{UN}, a data scientist responsible for prior analysis of the data.
%%
The third participant is \emph{TI}, a statistician with focus on epidemiology (8 years of experience), who assesses the statistical reliability of the tool and the underlying methods without a focus on a specific data set.
%%
\subsection{The Hepatic Steatosis Data Set}
We employ the data set used by Niemann et al.~\cite{Niemann2014} to identify predictive features w.r.t. the reversible hepatic steatosis disorder.
%%
The dichotomous target feature is derived from the liver fat concentration measured using MRI scans.
%%
Liver fat concentrations of no more than 10\% are mapped to the `negative' class; values greater than 10\% are mapped to the `positive' class to indicate absence or presence of the disease.
%%
The data set contains labels for 578 participants. %for which the MRI data was available to that point.
%%
The MRI scans for each subject are only available in SHIP-2. %, so the target feature is only present the last study wave.

%%
Apart from the target feature, the data set contains 199 features comprising sociodemographic features (e.g., gender, age), consumption behavior (e.g., alcohol and tobacco), laboratory data (e.g., sera concentrations), and two features depicting the liver ultrasound.
%%
The acquisition wave is denoted using the appendix; 85 features with appendix \emph{s0} denote their affiliation to SHIP-0 (first study moment), 50 features for \emph{s1} and 55 for \emph{s2}, alongside with 10 time-independent SNPs (DNA base pairs).
%%
Niemann et al.~\cite{Niemann2014} show different class distributions of liver fat concentrations of women and men.
%%
For women, an association between age and liver fat was identified.
%%
An appropriate cut-off value of 52 years, which is the approximate entry age for the menopause was set, yielding the most homogeneous class distribution within the resulting subsets.
%%
Based on these observations, we perform our analysis on three populations: \emph{males}, \emph{females (all ages)} and \emph{females older than 52 years}.
%%
\subsection{The Breast Density Data Set}
%%
The breast density data set was compiled to find associations between the parenchyma tissue proportion in the female breast compared to other features in the data.
%%
Breast density is denoted as the ratio between parenchyma and cellular connective tissue and has been shown to be associated with breast cancer.
%%
Studies describe a four to five times increased risk of getting breast cancer for participants with a breast density above 50\% \cite{Mccormack2006}.
